---
title: "Integrated Gamma Copula Definition"
output: 
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
vignette: >
  %\VignetteIndexEntry{Integrated Gamma Copula Definition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r}
library(igcop)
```

The generic copula class (DJ and interpolated DJ) is characterized by a _generating function_ $\psi:[0, \infty) \rightarrow (0, 1]$, which needs to be a concave distribution function or convex survival function.

This manuscript provides results for a generic copula generated by some $\psi$, in addition to the IG and IGL classes. Note that $\psi$ need not have an inverse -- the results still hold for the left-inverse, $\psi^{\leftarrow}$. 

# Preliminary Functions

## Incomplete Gamma Function

The (Upper) Incomplete Gamma function is defined as
$$\Gamma^{*}(\alpha, x) = \int_x^{\infty} t ^ {\alpha - 1} e ^ {-t} dt$$
for $\alpha > 0$ and $x \ge 0$, and the Gamma function is defined as $\Gamma(\alpha) = \Gamma^{*}(\alpha, 0)$.

In R, the Gamma function is accessible through the function `gamma()`, and the igcop package makes the Incomplete Gamma function accessible through the `igamma()` function. 

To make the `igamma()` function, the igcop package uses the Gamma distribution function `pgamma()` with `scale` or `rate` equal to 1, and `shape` equal to $\alpha$ as an indirect way of accessing the Incomplete Gamma function. Denoting this Gamma cdf as $F_{\alpha}$, the relationship is
$$F_{\alpha}(x) = 1 - \frac{\Gamma^{*}(\alpha, x)}{\Gamma(\alpha)}.$$

## Generating Function

The IG and IGL copula families result from a parametric generating function $\Psi_k$ for $k > 1$, defined below. 

| Function | Formula | igcop function |
|:--------:|:-------:|:--------------:|
| $\Psi_k(y)$ | $$
\begin{cases}
  y \frac{\Gamma(k) - \Gamma^{*}(k, y ^ {-1})}{\Gamma(k - 1)} +
    \frac{\Gamma^{*}(k - 1,y ^ {-1})}{\Gamma(k - 1)}, & y > 0;\\
  0, & y = 0
\end{cases}
$$ | `igl_gen()` |
| $\Psi_k'(y)$ | $$
\frac{\Gamma(k) - \Gamma^{*}(k, 1  / y)}
     {\Gamma(k - 1)}
$$ | `igl_Dgen()` |
| $\Psi_k''(y)$ | $$
-\frac{y ^ {-k - 1} \exp(-1 / y)}
      {\Gamma(k - 1)}
$$ | `igl_DDgen()` |

The "kappa" version of the generating function is also important. 

General definition.

| Function | Formula |
|:--------:|:-------:|
| $\kappa_{\psi}(y)$  | $\psi(y) - y \psi'(y)$ |
| $\kappa_{\psi}'(y)$ | $-y \psi''(y)$ |

Taking our specific generating function in place of $\psi$:

| Function | Formula | igcop function |
|:--------:|:-------:|:--------------:|
| $\kappa_k(y) := \kappa_{\Psi_k}(y)$ | $$
\frac{\Gamma^{*}(k - 1, 1 / y)}
     {\Gamma(k - 1)} = 1 - \text{pgamma}(1 / y, k - 1)
$$ | NA |
| $\kappa_k'(y)$ | $$
\frac{\exp(-1/y)}{y ^ k \Gamma(k - 1)} =
\frac{\text{dgamma}(1 / y, k - 1)}{y ^ 2}
$$ | NA | 



## Interpolating Function

The copula arising from the generating function $\psi$ gives rise to a parametric copula family governed by a parameter $\theta > 0$, where $\theta = 0$ is the independence copula, and $\theta \rightarrow \infty$ results in the original copula. As such, this copula family _interpolates_ a DJ copula, giving rise to an even wider class of copulas. 

The interpolated class of copulas relies on an _interpolating function_ function $H_\psi(\cdot; \theta): [1,\infty) \rightarrow (0, 1]$, defined below.

Generic definition of the interpolating function:

| Function | Formula |
|:--------:|:-------:|
| $H_{\psi}(y; \eta)$ | $$
\begin{cases}
  \frac{1}{y}\psi\left(\frac{1}{\eta \log y}\right), & y > 1;\\
  1, & y = 1
\end{cases}
$$ |
| $\text{D}_1 H_{\psi}(t; \eta)$ | $$
  -\frac{1}{t ^ 2}
  \left[
    \psi\left(\frac{1}{\eta \log t}\right) + 
    \frac{1}{\eta (\log t) ^ 2} \psi'\left(\frac{1}{\eta \log t}\right)
  \right]
$$ |
| $\text{D}_2 H_k(t; \eta)$ | $$
-\frac{1}{t \eta ^ 2 \log t}
\psi'\left(\frac{1}{\eta \log t}\right)
$$ |

Definition of the interpolating function using the IGL generating function $\Psi_k$ (taking $H_k := H_{\Psi_k}$):

| Function | Formula | igcop function |
|:--------:|:-------:|:--------------:|
| $H_k(t; \eta)$ | $$
\begin{cases}
  \frac{1}{t}\Psi_k\left(\frac{1}{\eta \log t}\right), & t > 1;\\
  1, & t = 1
\end{cases}
$$ | `ig_gen()` |
| $\text{D}_1 H_k(t; \eta)$ | $$
-\frac{
  (\log t + 1) (\Gamma(k) - \Gamma^{*}(k, \eta \log t)) +
  \eta (\log t) ^ 2 \Gamma^{*}(k - 1, \eta \log t)
}{
  \eta (t \log t) ^ 2 \Gamma(k - 1)
}
$$ | `ig_D1gen()` |
| $\text{D}_2 H_k(t; \eta)$ | $$
-\frac{1}{t \eta ^ 2 \log t}
\frac{\Gamma(k) - \Gamma^{*}(k,\eta \log t)}{\Gamma(k)}
$$ | NA |






For the IG copula, $H_k := H_{\Psi_k}$ (function `ig_gen()` in the igcop package). Note that $H_k(\cdot; \theta)$ is strictly decreasing
for all $\theta > 0$ and $k > 1$, so $H_k^{-1}(\cdot; \theta)$
is the unique inverse function of $H_k(\cdot; \theta)$.


# Copula Formulas

## DJ Copula Class

The Durante-Jaworski copula class:

| Quantity | Formula | 
|:--------:|:-------:|
| $C(u, v; \psi)$ | $$
u + v - 1 + (1 - u) \psi\left((1 - u) ^ {-1} \psi^{\leftarrow}(1 - v)\right)
$$ |
| $C_{2 | 1}(v | u; \psi)$ | $$
1 - \kappa_{\psi}\left((1 - u) ^ {-1} \psi ^ {\leftarrow}(1 - v) \right)
$$ |
| $C_{1 | 2}(u | v; \psi)$ | $$
1 - \frac{\psi'\left((1 - u) ^ {-1} \psi ^ {\leftarrow}(1 - v)\right)}
         {\psi'\left(\psi ^ {\leftarrow}(1 - v)\right)}
$$ |
| $c(u, v; \psi)$ | $$
-\frac{
  \psi ^ {\leftarrow}(1 - v) 
  \psi''\left((1 - u)^{-1} \psi^{\leftarrow}(1 - v)\right)
}{
  (1 - u) ^ 2 \psi'\left(\psi^{\leftarrow}(1 - v)\right)
}
$$ |
| $C_{2 | 1}^{-1}(\tau | u; \psi)$ | $$
1 - \psi((1 - u) \kappa_{\psi}^{\leftarrow}(1 - \tau))
$$ |
| $C_{1 | 2}^{-1}(\tau | v; \psi)$ |  |

Note that @coia in E.1.2 writes these distributional quantities in terms of the reflection copula. 

## DJ Copula Class with $\Psi_k$: IGL Copula Class

The parameter space is $k > 1$. **Note**: Values of $k < 2$ lead to unstable computations -- especially with $k$ close to 1. 

| Quantity | Formula | distplyr function |
|:--------:|:-------:|:-----------------:|
| $C(u, v; k)$ | $$
u + v - 1 + (1 - u) \Psi_k\left((1 - u) ^ {-1} \Psi_k^{-1}(1 - v)\right)
$$ | `piglcop()` |
| $C_{2 | 1}(v | u; k)$ | $$
1- \frac{\Gamma^{*}\left(k - 1, (1 - u) / 
           \Psi_k ^ {-1}(1 - v)\right)}
        {\Gamma\left(k-1\right)}
$$ | `pcondiglcop()`, `pcondiglcop21()` |
| $C_{1 | 2}(u | v; k)$ | $$
1 - \frac{\Gamma(k) - \Gamma^{*}\left(k, (1 - u) / 
            \Psi_k ^ {-1}(1 - v)\right)}
         {\Gamma(k) - \Gamma^{*}\left(k, 1 / 
            \Psi_k ^ {-1}(1 - v)\right)}
$$ | `pcondiglcop12()` |
| $c(u, v; k)$ | $$\frac{
  (1 - u)^{k-1}
  \left[\Psi_k^{-1}(1 - v)\right] ^ {-k}
  \exp\left\{-(1 - u) / \Psi_k^{-1}(1 - v)\right\} 
}{
  \Gamma(k) - \Gamma\left(k, 1 / \Psi_k^{-1}(1 - v)\right)
}$$ | `diglcop()` |
| $C_{2 | 1}^{-1}(\tau | u; k)$ | $$
1 - \Psi_k\left(\frac{1 - u}{\text{qgamma}(\tau, k - 1)}\right)
$$ | `qcondiglcop()`, `qcondiglcop21()` |
| $C_{1 | 2}^{-1}(\tau | v; k)$ |  | `qcondiglcop12()` |

## Interpolated DJ Copula Class



| Quantity | Formula | 
|:--------:|:-------:|
| $C(u, v; \theta, \psi)$ | $$
u + v - 1 + (1 - u) H_{\psi}\left(H_{\psi} ^ {\leftarrow}(1 - v; \theta); \theta (1 - u)\right)
$$ |
| $C_{2 | 1}(v | u; \theta, \psi)$ | $$
1 - H_{\kappa_{\psi}}\left(H_{\psi} ^ {\leftarrow}(1 - v; \theta); \theta (1 - u)\right)
$$ |
| $C_{1 | 2}(u | v; \theta, \psi)$ | $$
1 - (1 - u) \frac{
  \psi\left(\frac{1}{\theta (1 - u) \log t}\right) + 
  \frac{1}{\theta (1 - u) (\log t) ^ 2} 
  \psi'\left(\frac{1}{\theta (1 - u) \log t}\right)
} {
  \psi\left(\frac{1}{\theta \log t}\right) +
  \frac{1}{\theta (\log t) ^ 2} 
  \psi'\left(\frac{1}{\theta \log t}\right)
}
$$ where $t = H_{\psi} ^ {\leftarrow}(1 - v; \theta)$ |
| $c(u, v; \theta, \psi)$ | $$
\frac{
  \kappa_{\psi}\left(\frac{1}{\theta (1 - u) \log t}\right) +
  \frac{1}{\theta (1 - u) (\log t) ^ 2}
  \kappa_{\psi}'\left(\frac{1}{\theta (1 - u) \log t}\right)
} {
  \psi\left(\frac{1}{\theta \log t}\right) +
  \frac{1}{\theta (\log t) ^ 2}
  \psi'\left(\frac{1}{\theta \log t}\right)
}
$$ where $t = H_{\psi} ^ {\leftarrow}(1 - v; \theta)$ |
| $C_{2 | 1}^{-1}(\tau | u; \theta, \psi)$ | $$
1 - H_{\psi}\left(H_{\kappa_{\psi}}^{\leftarrow}(1 - \tau; \theta (1 - u)); \theta \right)
$$ |
| $C_{1 | 2}^{-1}(\tau | v; \theta, \psi)$ |  |

Note that @coia in E.2.2 writes these distributional quantities in terms of the reflection copula. 


## Interpolated DJ Copula Class with $\Psi_k$: IG copula

The parameter space is $\theta \ge 0$, $k > 1$. **Note**: Values of $k < 2$ lead to unstable computations -- especially with $k$ close to 1. 

| Quantity | Formula | distplyr function |
|:--------:|:-------:|:-----------------:|
| $C(u, v; \theta, k)$ | $$
u + v - 1 + (1 - u) H_k\left(H_k ^ {-1}(1 - v; \theta); \theta (1 - u)\right)
$$ | `pigcop()` |
| $C_{2 | 1}(v | u; \theta, k)$ | $$
1 - \frac{
  \Gamma^{*}\left(k - 1, \theta (1 - u) 
  \log\left(H_k^{-1}(1 - v; \theta)\right)\right)
}{
  H_k^{-1}(1 - v; \theta) \Gamma(k - 1)
}
$$ | `pcondigcop()`, ``pcondigcop21()`` |
| $C_{1 | 2}(u | v; \theta, k)$ | $$
1 - (1 - u) \frac{
  \text{D}_1 H_k\left(H_k^{-1}(1 - v; \theta); 
  \theta(1 - u)\right)
}{
  \text{D}_1 H_k\left(H_k^{-1}(1 - v; \theta); 
  \theta\right)
} 
$$ | `pcondigcop12()` |
| $c(u, v; \theta, k)$ | $$
-\frac{
  \theta ^ {k - 1} (1 - u) ^ {k-1} (\log t_v) ^ {k - 2} t_v ^ {-\theta (1 - u)} + 
  \Gamma^{*}(k - 1, \theta (1 - u) \log  t_v)
}{
  \Gamma(k - 1) t_v ^ 2 \text{D}_1 H_k(t_v; \theta)
}
$$ | `digcop()` |
| $C_{2 | 1}^{-1}(\tau | u; \theta, k)$ |  | `qcondigcop()`, `qcondigcop21()` |
| $C_{1 | 2}^{-1}(\tau | v; \theta, k)$ |  | `qcondigcop12()` |

(Note that @coia mistakenly writes $H_{\psi}$ instead of $H_k$ in the formula for $C_{\text{IG}, 1 | 2}$.)

The copula family is named "Integrated Gamma" because $\Psi_k'(1 / \cdot)$ is proportional to the Gamma distribution cdf with scale parameter 1 and shape parameter $k$. 

These two families are related, in that 
$$\lim_{\theta \rightarrow \infty} C_{IG}(u, v; \theta, k) = C_{IGL}(u, v; k)$$
for all $k$. Further, taking $\theta = 0$ results in the independence copula, so that the IG copula family is an "interpolation" between the independence copula and each IGL copula.






# Translating formulas to code

## Generating Function $\Psi_k$

Noting that $\Gamma(k - 1) = \Gamma(k) / (k - 1)$,

$$
\Psi_k(t) = 
y (k - 1) \left(1 - 
    \frac{\Gamma^{*}(k, y ^ {-1})}{\Gamma(k)}
    \right) +
    \left(\frac{\Gamma^{*}(k - 1,y ^ {-1})}{\Gamma(k - 1)}\right)
$$

The two major parentheses can be calculated from the Gamma cdf: the first set being `pgamma(1 / y, shape = k)`, and the second set being `pgamma(1 / y, shape = k - 1, lower.tail = FALSE)`. Here is the definition of `igl_gen`:

```{r}
igl_gen
```

The derivative is proportional to the Gamma density. Using the same Gamma function identity:

$$
\Psi_k'(y) = 
(k - 1) \left(1 - \frac{\Gamma^{*}(k, y ^ {-1})}{\Gamma(k)}\right)
$$



# `igcond` and `igcondinv` functions

The computation of the IGL 2|1 quantile function is simple. But, computing
the IGL 2|1 quantile function in a "direct" way first requires
computing an inverse. This section describes a way to avoid this "nested
inverse" with functions in the igcop package called `igcond`
and `igcondinv`.

The 2|1 distribution function of an IG copula is
$$C_{IGL,2|1}(v | u; \theta, k) 
= 1 - \frac{\bar{F}_{k-1}\left(\theta \bar{u} \log H_k ^ {-1}(\bar{v}; \theta)\right)}
           {H_k ^ {-1} (\bar{v}; \theta)}\\ 
= 1- \varphi_k\left(H_k ^ {-1}(\bar{v}; \theta); \theta \bar{u} \right),$$
where $\bar{u} = 1 - u$ and $\bar{v} = 1 - v$, and $\bar{F}_{k-1}\left(x\right)=\Gamma^{*}\left(k-1,x\right)/\Gamma\left(k-1\right)$
for $x \geq 0$ is the Gamma survival function with shape parameter
$k-1$ and unit scale parameter, and 
$$\varphi_k(x; \eta) = x^{-1} \bar{F}_{k-1}(\eta \log x)$$
for $x>1$. So, computing $C_{2|1}^{\leftarrow}$ amounts to computing
the inverse of $\varphi_k\left(\cdot;\theta\right)$:
$$C_{IGL, 2|1} ^ {\leftarrow}(\tau | u; \theta, k) 
= 1 - H_k\left(\varphi_k ^ {\leftarrow}(1 - \tau; \theta \bar{u}); \theta \right),$$
$\tau \in (0, 1)$. Without this $\varphi_k$ function,
one would first need to compute $H_k^{\leftarrow}(\bar{v}; \theta)$
to compute $C_{IGL, 2|1} ^ {\leftarrow}(v | u; \theta, k)$, and
this may lead to inaccuracies due to error amplification.

In the the igcop package package, `igcond` is $\varphi_k$,
and `igcondinv` is $\varphi_k^{\leftarrow}$. Here are some
plots of $\varphi_k$:

```{r plots, fig.width=5, fig.height=2, warning=FALSE, echo=FALSE} 
library(ggplot2)
phi <- function(theta, k) function(x) pgamma(theta*log(x), k-1, lower.tail=FALSE) / x
ggplot(data.frame(x=c(1,10)), aes(x)) +
   stat_function(fun=phi(2,1.3), mapping=aes(colour="theta=2")) +
   stat_function(fun=phi(5,1.3), mapping=aes(colour="theta=5")) +
   stat_function(fun=phi(10,1.3), mapping=aes(colour="theta=10")) +
   labs(title="k=1.3", y=expression(varphi[k])) +
   theme(axis.title.y=element_text(angle=0))
ggplot(data.frame(x=c(1,10)), aes(x)) +
   stat_function(fun=phi(2,3), mapping=aes(colour="theta=2")) +
   stat_function(fun=phi(5,3), mapping=aes(colour="theta=5")) +
   stat_function(fun=phi(10,3), mapping=aes(colour="theta=10")) +
   labs(title="k=3", y=expression(varphi[k])) +
   theme(axis.title.y=element_text(angle=0))
ggplot(data.frame(x=c(1,10)), aes(x)) +
   stat_function(fun=phi(2,10), mapping=aes(colour="theta=2")) +
   stat_function(fun=phi(5,10), mapping=aes(colour="theta=5")) +
   stat_function(fun=phi(10,10), mapping=aes(colour="theta=10")) +
   labs(title="k=10", y=expression(varphi[k])) +
   theme(axis.title.y=element_text(angle=0))
```

To compute `igcondinv` by solving $\varphi_k(x; \eta) = p$
for $x$, the Newton-Raphson algorithm is used to solve $g(x) = 0$
for $x$, where
\[
g\left(x\right)=xp-\bar{F}_{k-1}\left(\eta\log x\right),
\]
with derivative
\[
g'\left(x\right)=p+\frac{\eta}{x}f_{k-1}\left(\eta\log x\right).
\]
A starting value is used by noting that $\varphi_k$ is the product
of two invertible survival functions, and is therefore smaller than
those two survival functions. The smaller of the two *roots*
of these survival functions is therefore an upper bound for the root
of $\varphi_k$. The starting point is taken to be immediately to
the left of this upper bound.

\section{`ig_geninv` function{#sec:ig_geninv}}

Although computing $H_k^{\leftarrow}$ can be bypassed when computing
the IG 2|1 quantile function, it is needed when computing other copula
quantities, such as the copula distribution function. The inverse
$H_k^{\leftarrow}$ is called `ig_geninv` in the igcop package.

To compute $H_k^{\leftarrow}\left(p,\theta\right)$ for $p\in\left(0,1\right)$
and some $\theta>0$, Newton-Raphson can be used to solve $h\left(x\right)=0$
for $x$, where
\[
h\left(x\right)=xp-\Psi_k\left(\frac{1}{\theta\log x}\right).
\]
This function has derivative
\[
h'\left(x\right)=p+\frac{1}{\theta x\left(\log x\right)^2}\Psi_k'\left(\frac{1}{\theta\log x}\right).
\]
 

To get a good starting value for the algorithm, note that the function
$H_k$ is a product of two survival functions, just like $\varphi_k$
is (as described in Section~\@ref(sec:igcondinv)). And like inverting
$\varphi_k$, the inverse $H_k^{\leftarrow}\left(p;\theta\right)$
for some $\theta>0$ has an upper bound that's the minimum of the
roots of the individual survival functions:
\[
\min\left\{ \frac{1}{p},\exp\left(\frac{1}{\theta\Psi_k^{\leftarrow}\left(p\right)}\right)\right\} ,
\]
where $\Psi_k$ is defined in Equation~(\@ref(eq:cnstr-Psi)). Computing
the inverse $\Psi_k^{\leftarrow}$ should not be problematic in
terms of accuracy of the final result, since this computation is only
used to determine a starting value.
\end{document}
