---
title: "Formulas for the IG and IGL Copula Families"
output: 
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
vignette: >
  %\VignetteIndexEntry{Formulas for the IG and IGL Copula Families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ref.bib
---


```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(igcop)
library(tidyverse)
select_alpha <- c(0.1, 0.5, 1, 3, 9)
select_eta <- c(0.1, 2, 20, 200)
alpha_title <- fct_reorder(paste("alpha =", select_alpha), select_alpha)
eta_title <- fct_reorder(paste("eta =", select_eta), select_eta)
```

The formulas behind the Integrated Gamma (IG) and IG Limit (IGL) copula families are best presented in terms of their corresponding wider classes of copulas, since the formulas do not usually simplify. Specifically,

- the IGL copula family is a special case of a class of copulas first defined by Durante and Jaworski [@DJ] (which we call the _DJ copula class_), and
- the IG copula family is a special cases of an interpolated version of the DJ copula class [@coia2017].

# Formulas

## Definition via cdf

Both the DJ and the Interpolated DJ copula classes are characterized by a _generating function_ $\psi:[0, \infty) \rightarrow (0, 1]$, which needs to be a concave distribution function or convex survival function.

The DJ copula class (corresponding to the IGL copula family) has cdf
$$C_{\text{DJ}}(u, v; \psi) = u + v - 1 + (1 - u) \psi\left((1 - u) \psi^{\leftarrow}(1 - v)\right),$$
for $(u, v)$ in the unit square, and $\psi^{\leftarrow}$ the left-inverse of $\psi$. 

But, this class does not contain the independence copula, so we can introduce an interpolating parameter $\theta \geq 0$ such that $\theta = 0$ results in the independence copula, and $\theta = \infty$ results in the DJ copula class -- hence _interpolating_ DJ copulas with the independence copula. The cdf of an interpolated DJ copula (corresponding to the IG copula family) is
$$C_{\text{intDJ}}(u, v; \theta, \psi) = u + v - 1 + (1 - u) H_{\psi}\left(H_{\psi}^{\leftarrow}(1 - v; \theta); (1 - u) \theta \right)$$
for $(u, v)$ in the unit square, where $H_{\psi}$ is the interpolating function of $\psi$, defined as
$$H_{\psi}(x; \eta) = e ^ {-x} \psi(\eta x)$$
for $x > 0$ and $\eta \geq 0$. The derivative of the interpolating function (with respect to the first argument) shows up in formulas for copula quantities, and is
$$\text{D}_1 H_{\psi}(x; \eta) = - e ^ {-x} (\psi(\eta x) - \eta \psi'(\eta x)).$$

Although the DJ copula class can be considered part of the Interpolated DJ class if we include $\theta = \infty$ in the parameter space, the formulas do not simplify in an obvious way, so I'll continue to give the two classes separate treatment. 

These formulas are originally found in [@coia2017], but under a different (and more complex) parameterization: the $\psi$ function is $\psi(1/x)$ and the $H$ function is $H_{\psi}(x; \theta) = \frac{1}{x} \psi(1 / (\theta \log x))$, and the copula formulas differ slightly.

## Kappa Transform of $\psi$

Besides $H_{\psi}$, another transformation of $\psi$ is necessary for writing formulas succinctly:
$$\kappa_{\psi}(x) = \frac{\text{d}}{\text{d}x} x \psi(x) = \psi(x) + x \psi'(x)$$
for $x > 0$.

By the way, not all choices of $\psi$ result in valid DJ / Interpolated DJ copulas -- the requirement has to do with the $\kappa$ function. So, in case you want to go in the opposite direction, and obtain $\psi$ from a choice of $\kappa$, we can do so by solving the differential equation in the definition of $\kappa$, to get
$$\psi(x) = \frac{1}{x} \int_{0}^{x} \kappa(t) \text{d}t.$$

## Copula Quantities: DJ copula class (IGL copula family)

To simplify formulas, denote $y = \psi^{\leftarrow}(1 - v)$. Throughout, $p$, $u$, and $v$ are numbers between 0 and 1. 

The density of a DJ copula (and thus an IGL copula) is
$$c_{\text{DJ}}(u, v; \psi) = (1 - u) \frac{\kappa_{\psi}'((1 - u) y)}{\psi'(y)}$$

The two families of conditional distributions, obtained by conditioning on either the 1st or 2nd variable, are
\begin{equation}
\begin{split}
C_{\text{DJ}, 2 | 1}(v | u; \psi) & = 1 - \kappa_{\psi}((1 - u) y) \\
C_{\text{DJ}, 1 | 2}(u | v; \psi) & = 1 - (1 - u) ^ 2 \frac{\psi'((1 - u) y)}{\psi'(y)}.
\end{split}
\end{equation}
The corresponding quantile functions are the (left) inverses of these functions, for which the 2|1 quantile function has a closed form:
$$C_{\text{DJ}, 2 | 1}^{-1}(p | u; \psi) = 1 - \psi\left((1 - u) ^ {-1} \kappa_{\psi}^{\leftarrow}(1 - p) \right)$$

To check these equations, note that $\frac{\text{d}y}{\text{d}v} = - 1 / \psi'(y)$. Note that [@coia2017] in E.1.2 mistakenly writes these distributional quantities as if the family was defined as its reflections.

## Copula Quantities: Interpolated DJ copula class (IG copula family)

To simplify formulas, denote $y = H_{\psi} ^ {\leftarrow}(1 - v; \theta)$. Throughout, $p$, $u$, and $v$ are numbers between 0 and 1. 

The density of an interpolated DJ copula (and thus an IG copula) is
\begin{equation}
c_{\text{intDJ}}(u, v; \theta, \psi) = 
  \frac{\text{D}_1 H_{\kappa_{\psi}}(y; (1 - u) \theta)}
       {\text{D}_1 H_{\psi}(y; \theta)}
\end{equation}

The two families of conditional distributions, obtained by conditioning on either the 1st or 2nd variable, have a convenient form if we consider the interpolating function $H$ _of the $\kappa$ function_ instead of the $\psi$ function:
\begin{equation}
\begin{split}
C_{\text{intDJ}, 2 | 1}(v | u; \theta, \psi) & = 1 - H_{\kappa_{\psi}}\left(y; (1 - u) \theta \right) \\
C_{\text{intDJ}, 1 | 2}(u | v; \theta, \psi) & = 1 - (1 - u) 
  \frac{\text{D}_1 H_{\psi}(y; (1 - u) \theta)}
       {\text{D}_1 H_{\psi}(y; \theta)}\\
\end{split}
\end{equation}
The corresponding quantile functions are the (left) inverses of these functions, for which the 2|1 quantile function has a closed form:
\begin{equation}
\begin{split}
C_{\text{intDJ}, 2 | 1}^{-1}(p | u; \theta, \psi) & = 1 - H_{\psi} \left(
  H_{\kappa_{\psi}} ^ {\leftarrow} (1 - p; (1 - u) \theta); \theta
\right)
\end{split}
\end{equation}

To check these equations, note that $\frac{\text{d}y}{\text{d}v} = - 1 / \text{D}_1 H_\psi(y; \theta)$.

# Formulas specific to IG and IGL Copulas

## Generating Functions

The generating functions of the IG and IGL copula families rely on the Gamma distribution, which has cdf
$$F_{\alpha}(x) = \frac{\Gamma(\alpha) - \Gamma^{*}(\alpha, x)}{\Gamma(\alpha)}$$
for $\alpha > 0$ and $x \ge 0$,
where $\Gamma$ is the Gamma function, and $\Gamma^{*}$ is the (upper) incomplete Gamma function defined as
$$\Gamma^{*}(\alpha, x) = \int_x^{\infty} t ^ {\alpha - 1} e ^ {-t} \text{d} t.$$
The density function is denoted by $f_{\alpha}$.

To emphasize the dependence of $\psi$, $\kappa$, and $H$ on the parameter $\alpha$, the parameter is written as a subscript. The $\psi$ function and its derivative are
\begin{equation}
\begin{split}
\psi_{\alpha}(x) & = 1 - F_{\alpha}(x) + \frac{\alpha}{x} F_{\alpha + 1}(x); \\
\psi_{\alpha}'(x) & = - \frac{\alpha}{x ^ 2} F_{\alpha + 1}(x).
\end{split}
\end{equation}
Here are some plots of these functions.
```{r, echo = FALSE, fig.width = 7, fig.height = 2, fig.align = "center"}
expand_grid(nesting(alpha = select_alpha,
                    alpha_title = alpha_title),
            x = seq(0, 8, length.out = 1000)) %>% 
  mutate(psi = map2_dbl(x, alpha, igcop:::igl_gen)) %>% 
  ggplot(aes(x, psi)) +
  facet_wrap(~ alpha_title, nrow = 1) +
  geom_line() +
  theme_bw() +
  ylab(expression(psi[alpha](x))) +
  theme(axis.title.y = element_text(angle = 0))
```

The $\kappa$ function is the Gamma survival function,
$$\kappa_{\alpha}(x) = 1 - F_{\alpha}(x).$$

The $H$ function does not simplify, although for convenience, $H_{\psi_{\alpha}}$ is simply denoted $H_{\alpha}$. Here are some plots of these functions, compared with the plot of the negative exponential $e^{-x}$ as the faded dotted line. Notice that increasing $\alpha$ draws $H_{\alpha}(\cdot, \eta)$ closer to the negative exponential, whereas increasing $\eta$ pulls it further.

```{r, echo = FALSE, fig.width = 7, fig.height = 4, fig.align = "center"}
df <- expand_grid(nesting(alpha = select_alpha,
                    alpha_title = alpha_title),
            nesting(eta = select_eta,
                    eta_title = eta_title),
            x = seq(0, 2.5, length.out = 1000)) %>%
  mutate(H = pmap_dbl(list(x, eta, alpha), igcop:::interp_gen),
         e = exp(-x)) %>%
  pivot_longer(cols = c("H", "e"), names_to = "fun", values_to = "eval")
ggplot(df, aes(x, eval)) +
  facet_grid(eta_title ~ alpha_title) +
  geom_line(aes(group = fun, linetype = fun, alpha = fun)) +
  ylab(expression(H[alpha](x*";"~eta))) +
  guides(linetype = FALSE, alpha = FALSE) + 
  scale_alpha_manual(values = c(0.5, 1)) +
  scale_linetype_manual(values = c("dotted", "solid")) +
  scale_x_continuous(breaks = 0:2) +
  theme_bw() +
  theme(axis.title.y = element_text(angle = 0))
```


## Copula Quantities

Most of the formulas for copula quantities do not simplify nicely from their general forms. But, here are the ones that do.

For the IGL copula family:
\begin{equation}
\begin{split}
C_{\text{IGL}, 1 | 2}(u | v; \theta, \alpha) & = 
  1 - \frac{F_{\alpha + 1}((1 - u) y)}{F_{\alpha + 1}(y)} \\
C_{\text{IGL}, 1 | 2}^{-1}(p | v; \theta, \alpha) & = 1 - \frac{1}{y}
  F_{\alpha + 1}^{-1}\left((1 - p) F_{\alpha + 1}(y)\right)
\end{split}
\end{equation}


# References
